Evolution of ORM Systems: Formal Foundations

### 2013-01-09 doc/description.tex
Improvements in data transformations

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete values of concrete columns. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i \lhd Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\subsection{Database Transformation}
The database is composed from two separated parts - database schema and data. Several transformation are defined to make manipulating a database easily in this paper.
\paragraph{Add Column}
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c)}{columns(t) \cup c}\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}
\paragraph{COPY COLUMN}
\begin{align*}
	\mathbf{copyColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c }
	\end{cases} \\
	l_s,l_t,l_c \in Label,d \lhd Database
\end{align*}
\paragraph{Add Table}
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e)}{tables(d) \cup c}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}
\paragraph{Insert row}
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) }{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}
\paragraph{Insert Values}
\begin{align*}
	\mathbf{insertValues} : Label \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(l, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(l, head(p), d)\bullet insertValues(l, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r_t) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r_t), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
l \in Label, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{COPY VALUES}
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Instancece \times Instance) \times Database \times Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
 		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc} }{\begin{gathered}
			\forall p \in f_m(t_s, t_t)  : \\ insertValues(name(second(p)), \\ [l_{tc},value(pairByRef(l_{sc}, first(p))),d)
		\end{gathered}}
 	\end{cases}
\end{align*}

@@

The copy property can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, where the $copyData$ transformation is used.
\begin{align*}
	\mathbf{CopyProperty} : Label \times Label \times Label \times (Instance \times Instance) \times State \rightarrow State \\
	\Psi(CopyProperty(l_s, l_t, l_p, f_m), s) = \inference{\exists c \in classes(s) : name(c) = l_s \\ \wedge \exists p \in properties(c) : name(p) = l_p }{application(s) \circ addProp(l_t, p)} \\\\
	\Phi(CopyProperty(l_s, l_t, l_p, f_m), s) = \begin{cases}
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) \leq 1 }{\begin{gathered}
			database(s) \bullet copyColumn(l_s, l_t, l_p) \\ \bullet copyValues(l_s, l_t, l_p, f_m) 
			\end{gathered}} \\ \\
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) > 1}{\begin{gathered}
			database(s) \bullet copyTable(l_s, l_t, l_p) \\ \bullet copyRows(l_s, l_t, l_p, f_m) 
			\end{gathered}}
 \end{cases}
\end{align*}

>>>>

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete values of concrete columns. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i \lhd Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \in Rows, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\subsection{Database Transformation}
The database is composed from two separated parts - database schema and data. Several transformation are defined to make manipulating a database easily.
\paragraph{Add Column} A new column can be add into a table if there are no name collisions. If the column has the $NOTNULL$ constraint, then the default value has to be added to all rows in the table.
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \not \in constraints(c) }{columns(t) \cup c}\\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \in constraints(c) }{columns(t) \cup c}\\ \bullet appendValues(l, [name(c), defVal(c)]) \\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}

\paragraph{Copy Column} The transformation creates a copy of a column in a another table in a database. The transformation changes only the structure of database - stored data are not changed.
\begin{align*}
	\mathbf{copyColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c} 
	\end{cases} \\
	l_s,l_t,l_c \in Label \wedge l_s \neq l_t,d \lhd Database
\end{align*}
\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge foreignKeys(t) = \varnothing }{tables(d) \cup t}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}
\paragraph{Insert row} The function allow to insert a row(s) into a database.
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) \\ \wedge \forall e \in tableData(t, d) : pkValue(e) \neq pkValue(r)}{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}
\paragraph{Append Values} The function appends a pair to all rows of a table.
\begin{align*}
	\mathbf{appendValues} : Label \times Pair \times Database \rightarrow Database \\
	appemdValues(l, p, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l, \\ r = tableData(l, d)}{insertValue(r, p, d)} \\ \\
 		Inconsistent \; d
 	\end{cases}\\ \\
 	l \in Label, d \lhd Database
\end{align*}
\paragraph{Insert Value} The function inserts a new pair into an existing row.
\begin{align*}
	\mathbf{insertValues} : Row \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(r, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(r, head(p), d)\bullet insertValues(r, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
r \lhd Row, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{COPY VALUES}
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Instancece \times Instance) \times Database \times Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
 		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc} }{\begin{gathered}
			\forall p \in f_m(t_s, t_t)  : \\ insertValues(name(second(p)), \\ [l_{tc},value(pairByRef(l_{sc}, first(p))),d)
		\end{gathered}}
 	\end{cases}
\end{align*}

@@

The copy property can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, where the $copyData$ transformation is used.
\begin{align*}
	\mathbf{CopyProperty} : Label \times Label \times Label \times (Row \times Row) \times State \rightarrow State \\
	\Psi(CopyProperty(l_s, l_t, l_p, f_m), s) = \inference{\exists c \in classes(s) : name(c) = l_s \\ \wedge \exists p \in properties(c) : name(p) = l_p }{application(s) \circ addProp(l_t, p)} \\\\
	\Phi(CopyProperty(l_s, l_t, l_p, f_m), s) = \begin{cases}
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) \leq 1 }{\begin{gathered}
			database(s) \bullet copyColumn(l_s, l_t, l_p) \\ \bullet copyValues(l_s, l_t, l_p, f_m) 
			\end{gathered}} \\ \\
		\inference{\exists p \in properties(c), c \in classes(s) : \\ name(p) = l_p \wedge cardinality(p) > 1}{\begin{gathered}
			database(s) \bullet copyTable(l_s, l_t, l_p) \\ \bullet copyRows(l_s, l_t, l_p, f_m) 
			\end{gathered}}
 \end{cases}
\end{align*}
